Your time series is (probably) too short

### 2017-06-03 writing/manuscript_draft.Rmd
added to methods and intro of paper

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
fontsize: 11pt
geometry: margin=0.75in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \usepackage{setspace}             
    \usepackage{lineno}
    \usepackage[round]{natbib}
    \bibpunct[; ]{(}{)}{,}{a}{}{;}
    \linenumbers
---

@@

## Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. In addition, census data is often expensive, requiring a lot of people-hours or equipment. However, little work has actually addressed the length of time series required. In other words, when is a time series of census data long enough to address a question of interest? Here, I explore two approaches to address this question: one simulation-based and an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using simple simulations, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size. In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. Here I show 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length. These results point to the importance of sampling populations over long periods of time. Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance. Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.

## Introduction

@@

Observational studies and population censuses have become a cornerstone of modern ecology and conservation biology [@Magurran2010]. Longterm data is necessary to understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time. There are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually necessary? This question has important scientific and management implications. Scientifically, we need to know how long a time series is required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability. From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not. For instance the International Union for the Conservation of Nature (IUCN) suggest...In addition, because sampling is typically expensive, we do not want to over sample [@Gerber1999].

In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. Formally, statistical power is one minus the probability of a type II error, or false negative. In other words, it is a measure of how well you could detect a trend given it is actually significant. Prior to an experiment, one could set appropriate levels of power, $\alpha$, and the effect size to estimate the sample size required for the experiment. This approach, however, is not as simple for a time series as the data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding power calculations, simulation approaches can be used [@Gerrodette1987]. Simple population models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then one can run simulations for time series of different lengths to determine the minimum time series length required to be confident that you will be able to answer your question. In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power...

Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..
- need to expand this paragraph in line with what I wrote in the discussion...

@@

One approach to determining the minimum time series needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015]. We only briefly discuss this approach as it has been desribed elsewhere (citations). Essentially, we take a stochastic population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power. This approach requires us to determine values for our model parameters. As an example, we can take the following population model for population size $N$ at time $t$:
$$
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
$$
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of increase or decrease. Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated. This model represents linear growth with autocorrelation and noise. However, it is certainly possible to use any other population model.

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here statistical power is simply the number of trials that are themselves significant, as the real trend is different from zero. Then the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold. Historically, 0.8....s

In figure X, a number of simulated time series are shown for a set number of time periods (t=40). It is clear that statistical power increases quickly with increases in $t$ (Fig. Xb). Once power is greater than XX (the dotted line), that is the minimum time required ($t_{min}$

<!--
![optional caption text](example_theoretical_output.pdf)
-->

```{r,echo=F,fig.cap='theoretical approach\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

@@

```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
 setwd("~/Desktop/Research/time-series-project")
#Load keith database
load('data/Keith_et_al_2015_data.Rdata')

#Load our cleaned data
load('cleaned-data/pop_info_example_goodNAs.Rdata')
```

### Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with `r nrow(pop_info)` time. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required to sample according to IUCN standards under rule ... (citation needed).

@@

Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest. As we describe below, the minimum time series length calculation is particular to each question. Intuitively, we might expect a longer minimum time series length required given more complicated questions.

We assume that each time series in our database are long enough to include all necessary information (e.g. variability) about the population. In other words, they are a Representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and so forth. In this paper, we will try to ask a simple, specific question. We seek to determine the $T_{min}$ required to be confident in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power. Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which here we will assume is 0.8 by convention (citation).

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.

In the appendix, we explore other potential questions of interest...

@@

## Results

- summary of findings
- theoretical approach
  - example plot
- empirical approach
  - plots: distribution of min time, what min time correlations with
  - walk through one empirical example (maybe in appendix)

@@

We constructed a general population model where the trend over time could be a model parameter. We chose a simple model, but any other population model could be used. The specific model choice should be tailored to the population of interest, based on species ecology. We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8). We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/pop_info_example_goodNAs.Rdata")

@@

```{r poisson_model, echo=F}
#Poisson regression
pop_info$abs_overall_trend = abs(pop_info$overall_trend)
poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
#plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')

@@

```{r,echo=F,fig.cap='duh\\label{fig:correlates}'}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)

       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r mean(pop_info$min_time_for_power)` ($\sigma$=`r (var(pop_info$min_time_for_power))^0.5`), with a wide distribution (Fig. \ref{fig:min_time_dist}). As expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model here). Combined trend strength, population variance, and autocorrelation account for about `r 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)` of the variance in minimum time series length required (as proposed by CITE Zuur).

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None (need to do regression) of these explanatory variables were predictive of the minimum time series length required. Some differences if $t_{min}$ did occur between biological classes, but this can be attributed to the amount populations are know to fluctuate.

>>>>

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
fontsize: 12pt
geometry: margin=1in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \usepackage{setspace}             
    \usepackage{lineno}
    \usepackage[round]{natbib}
    \bibpunct[; ]{(}{)}{,}{a}{}{;}
    \linenumbers
---

@@

## Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. In addition, census data is often expensive, requiring a lot of people-hours or equipment. However, little work has actually addressed the length of time series required. In other words, when is a time series of census data long enough to address a question of interest? Here, I explore two approaches to address this question: one simulation-based and an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using simple simulations, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size. In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. Here I show 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length. These results point to the importance of sampling populations over long periods of time. Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance. Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

## Introduction - need to clean up and reduce text

@@

Observational studies and population censuses have become a cornerstone of modern ecology and conservation biology [@Magurran2010]. Longterm data is necessary to understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time. However, there are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually necessary? This question has important scientific and management implications. Scientifically, we need to know how long a time series is required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not. For example, the IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. In addition, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].

In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative. In other words, it is a measure of how well you could detect a trend given it is actually significant. Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple population models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite). Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time. However, most work has either focused on theoretical approaches or by examined a single population. As an example of the former, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work (citation), have focused either on theoretical aspects of monitoring design or focused on single species.

@@

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015] (and cite Bolker). We only briefly discuss this approach as it has been desribed elsewhere (citations). Essentially, we take a stochastic population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. growth rate). As an example, we can take the following population model for population size $N$ at time $t$:
$$
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
$$
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of increase or decrease. <!--Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated. This model represents linear growth with autocorrelation and noise. However, it is certainly possible to use any other population model.-->

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here statistical power is simply the fraction of simulations with significant trends in abundance from simple linear regression. Then, the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold (historically at 0.8).

In figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods (t=40). It is clear that statistical power increases quickly with increases in $t$ (Fig. \ref{fig:theoretical_approach}b). Once power is greater than XX (the dotted line), that is the minimum time required ($t_{min}$

```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power.\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

@@

```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
  setwd("~/Desktop/Research/time-series-project")

#Load keith et al. database (from GPDD database)
  load('data/Keith_et_al_2015_data.Rdata')

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/pop_info_example_goodNAs.Rdata')
```

### Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with `r nrow(pop_info)` time. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

@@

Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest. As we describe below, the minimum time series length calculation is particular to each question. Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, they are a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continueing until 1 possible contiguous subsample of length 40. In this paper, we will try to ask a simple, specific question. We seek to determine the $T_{min}$ required to be confident in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend. Complementary, there would be a 0.2 probability of a false negative.

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk. Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.

@@

## Results
<!--
- summary of findings
- theoretical approach
  - example plot
- empirical approach
  - plots: distribution of min time, what min time correlations with
  - walk through one empirical example (maybe in appendix)
-->

@@

We constructed a general population model where the trend over time could be a model parameter. We chose a simple model, but any other population model could be used. The specific model choice should be tailored to the population of interest, based on its ecology. We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8). We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/pop_info_example_goodNAs.Rdata")

@@

```{r poisson_model, echo=F,include=F}
#Poisson regression
pop_info$abs_overall_trend = abs(pop_info$overall_trend)
poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
#plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')

@@

```{r,echo=F,fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
  mtext('(a)',3,line=-1.5,at=0.033,cex=1.2,font = 2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.3)
    mtext('(b)',3,line=-1.5,at=0.128,cex=1.2,font = 2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)
       mtext('(c)',3,line=-1.5,at=0.98,cex=1.2,font = 2)
       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model from supp mat here). Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` percent of the variance in minimum time series length required (as proposed by CITE Zuur).

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None (need to do regression) of these explanatory variables were predictive of the minimum time series length required (supp mat). Some differences if $t_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e. the variability in population size).
